The Legacy of the European Post-Master in Urbanism at TU Delft: A Text Mining Approach

### 2023-07-05 analysis/paper/paper.qmd
Update analysis/paper/paper.qmd

# Load packages ----
## List packages to be loaded
packages <- c(
  "here",                               # Managing paths
  "tidyverse", "reshape2", "widyr",     # Data manipulation and visualisation
  "furrr",                              # Parallel use of purrr::map() functions
  "tidytext", "SnowballC", "tm",        # Text processing and analysis
  "topicmodels", "LDAvis",              # Topic modeling
  "tidygeocoder",                       # Working with spatial data
  "igraph", "ggraph"                    # Network analysis and visualization
)

@@

```{r get-pdf-text}
#| echo: false
#| cache: true

data_path <- here("analysis", "data", "derived_data", "emu_theses_with_text_and_location.csv")

if (!file.exists(data_path)) {
  # On the first run, this function might take extra time as a Python environment needs to be set up
  emu_theses$text <- convert_pdf_text(pdf_paths)

  # Write thesis dataframe to file
  write_csv(emu_theses, data_path)
} else {
  emu_theses$text <- read_csv(data_path)$text
  emu_theses$longitude <- read_csv(data_path)$longitude
  emu_theses$latitude <- read_csv(data_path)$latitude
}

@@

The imported dataset contains the text data in the `text` column and related metadata in the other columns...

@@

### Pre-processing

@@

For analyses in which word-level tokens were used, stop words, i.e., common words or words that do not add much meaning, were removed in the pre-processing stage. Global, subject-specific and document stop words [@hvitfeldt2021a] were differentiated. As all theses were written in English, the pre-made "snowball" lexicon was used to remove global stop words. <!-- We tried to create a custom global stop word list on the entire corpus, by removing high-frequency words, starting with 20 words and increasing by 10 until reaching words that are not appropriate as stop words, but no significant improvement was observed in the result. --> We manually constructed a list of subject-specific stop words based on our domain knowledge. Words that are commonly used to structure a theses, such as "preface", "contents", and "introduction", were added as subject-specific stop words as well. Finally, titles and author names were removed as document stop words. In the next pre-processing step, we normalised the resulting words, that is, reduced them to their canonical dictionary forms, the resulting words were lemmatised [@hvitfeldt2021a]. Lemmatisation was preferred over stemming as the former tends to be more meaningful and less degrading than the latter.

@@

# Add cleaned text column with removed stop words
emu_theses$text_clean <- emu_theses$text_clean |>
  str_remove_all(find_meta_stopwords(emu_theses)) |> 
  str_remove_all(urbanism_stopwords(
    add_stopwords = c("emu", "european postgraduate master in urbanism", "tu delft", "ku leuven", "upcs barcelona", "iuav venice"))) |>
  str_remove_all(thesis_stopwords(add_stopwords = c("advisor", "prof")))

# Unnest words and remove words of one and two letters
emu_theses_words <- emu_theses |> 
  select(title, text_clean) |> 
  unnest_tokens(output = word, input = text_clean) |>  # remove punctuation, convert to lower-case, separate all words
  anti_join(tidytext::stop_words, by = "word") |>  # remove stop words
  mutate(word = textstem::lemmatize_words(word))  # lemmatise words

@@

Finally, additional checks were made to the document in search for previously undetected stop words. Short words of one and two letters could be safely removed, while three-letter words were kept, as many of those, including "map" and "low", were meaningful and occurring in high frequency.

```{r short-stop-words}
#| dependson: stop-words

#short_words(emu_theses_words$word, l_word = 3, l_list = 100)

emu_theses_words <- emu_theses_words |>  
  filter(nchar(word) >= 3)  # keep only words that are at least three letters long

@@

@@

Similar to clustering on numeric data, topic modeling is a method of unsupervised classification of topics found in a group of documents [@silge2017]. We will use the Latent Dirichlet Allocation (LDA) algorithm for topic modeling. In LDA, each document is a mixture of topics and each topic is a mixture of words. The probability of a term being part of a topic is given by the $\beta$ (beta) statistic, while the probability of a document being part of a topic is given by the $\gamma$ (gamma) statistic.

```{r topic-model-lda}

# Prepare document-term matrix as input for LDA model
emu_dtm <- convert_to_dtm(emu_theses_words, "title", "word")

# Create LDA model
emu_lda <- LDA(emu_dtm, k = 2, control = list(seed = 2023))

```{r topic-model-beta}

# Extract topics as mixtures of terms from LDA model
emu_topics <- tidy(emu_lda, matrix = "beta")

# The top n most common words in each topic
emu_theses_top_terms <- get_top_words_per_topic(emu_topics, 20)

# TODO create visualisation function
emu_theses_top_terms |> 
  ggplot(aes(reorder(term, beta), beta, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic, scales = "free") +
  coord_flip()

# TODO can this be generalised to more than two topics?
beta_spread <- emu_theses_topics |> 
  mutate(topic = paste0("topic", topic)) |> 
  spread(topic, beta) |> 
  filter(topic1 > 0.001, topic2 > .001) |> 
  mutate(log_ratio = log2(topic2/topic1))

beta_spread %>%
  group_by(direction = log_ratio > 0) %>%
  slice_max(abs(log_ratio), n = 20) %>% 
  ungroup() %>%
  mutate(term = reorder(term, log_ratio)) %>%
  ggplot(aes(log_ratio, term)) +
  geom_col() +
  labs(x = "Log2 ratio of beta in topic 2 / topic 1", y = NULL)
```

```{r topic-model-gamma}

# Extract documents as mixtures of topics from LDA model
emu_documents <- tidy(emu_lda, matrix = "gamma")

emu_documents |> 
  ggplot(aes(x = reorder(str_sub(document, 1, 50), gamma), y = gamma, fill = topic)) + 
  geom_bar(stat = "identity") + 
  coord_flip()

emu_documents |> 
  inner_join(emu_theses |> 
               select(title, graduation_year), 
             by = c("document" = "title")) |> 
  ggplot(aes(x = reorder(str_sub(document, 1, 50), gamma), y = gamma, fill = topic)) + 
  geom_bar(stat = "identity") + 
  coord_flip() + 
  facet_wrap(~ graduation_year, ncol = 1, scales = "free", drop = TRUE)

emu_documents |> 
  inner_join(emu_theses |> 
               select(title, graduation_year), 
             by = c("document" = "title")) |> 
  ggplot(aes(x = factor(graduation_year), y = topic, fill = factor(topic))) +
  geom_col()
```

@@

```{r top-words-document}
# Which are the top n most frequently used words in each thesis?
get_top_words_per_document(data = emu_theses_words, top_n = 3)
```

```{r top-words-corpus}
# Which are the top n most frequently used words in all theses?
# TODO add `topic` to `fill` aesthetic
get_top_words_per_corpus(data = emu_theses_words, top_n = 50) |>
  ggplot(aes(reorder(word, n), n)) +
  geom_col(fill = "orange") +
  geom_text(aes(label = n), size = 2, hjust = 1.1) +
  coord_flip() +
  xlab("Word") +
  ylab("Frequency") +
  labs(title = paste0("Top ", n, " most used words in the corpus of theses")) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

@@

## Bigrams
emu_theses_bigrams <- emu_theses |> 
  get_ngrams(n = 2, title_col = "title", text_col = "text_clean", stem = TRUE)

spacyr::spacy_parse(emu_theses_bigrams$w_1)
data.frame(emu_theses_bigrams$w_1, textstem::lemmatize_words(emu_theses_bigrams$w_1))

set.seed(12345)

# with spacy
extract_rand_window <- function(x, p){
    first_index = sample(seq(length(x) - p + 1), 1)
    x[first_index:(first_index + p -1)]
    }

example_lemma <-list()
for (i in 1:10) {
  example_lemma[[i]] <- extract_rand_window(emu_theses_bigrams$w_1,10)
}

Original <- unlist(example_lemma)
Lemmatized <- spacyr::spacy_parse(Original)

# with textstem
Lemmatized2 <- textstem::lemmatize_words(Original)

lemma_example <- data.frame(Original, Lemmatized2)

lemma_example

data.frame(Lemmatized$token, Lemmatized$lemma, Lemmatized2)

bigram_counts <- emu_theses_bigrams |> 
  count(w_1, w_2, sort = TRUE)

bigram_stem_counts <- emu_theses_bigrams |> 
  count(w_1_stem, w_2_stem, sort = TRUE)

bigram_counts
bigram_stem_counts

## Trigrams
emu_theses_trigrams <- emu_theses |> 
  get_ngrams(n = 3, title_col = "title", text_col = "text")

trigram_counts <- emu_theses_trigrams |> 
  count(w_1, w_2, w_3, sort = TRUE)

@@

emu_theses_bigrams_top <- emu_theses_bigrams |>
  group_by(title) |>
  count(ngram, sort = TRUE) |>
  slice_max(n, n = 20) |>        ### top_n is superseded; better use slice_max instead
  # filter(n > 50)                ### filter words with a minimum count
  ungroup() |>
  mutate(bigram = reorder(ngram, n))

emu_theses_bigrams_top_all <- emu_theses_bigrams |>
  count(ngram, sort = TRUE) |>
  slice_max(n, n = 20) |>        ### top_n is superseded; better use slice_max instead
  # filter(n > 50)                ### filter words with a minimum count
  mutate(ngram = reorder(ngram, n))

emu_theses_bigrams_top |>
  filter(title == unique(title)[1:10]) |> 
  ggplot(aes(x = ngram, y = n, fill = title)) +
  geom_col() +
  xlab(NULL) +
  coord_flip() +
  labs(y = "Count",
       x = "Unique words",
       title = "Count of unique words found in thesis",
       subtitle = "Stop words were removed from the list") +
  facet_wrap( ~ title, scales = "free", ncol = 2) +
  theme(legend.position = "none") 
```

The `tf-idf` statistic shows words that are the most important to one document in a collection of documents [@silge2017]. In this case, it shows high values for the names of thesis locations that tend to be high-frequency thesis-specific words.

```{r tf-idf}

## Calculate tf-idf statistic
emu_theses_tf_idf <- emu_theses_words |> 
  get_tf_idf(title_col = "title", word_col = "word")

## Get top n words most specific to each thesis 
emu_theses_tf_idf |> 
  group_by(title) |> 
  top_n(5, tf_idf) |> 
  arrange(desc(tf_idf), .by_group = TRUE)

```{r}
# Preview term frequency distributions in a subset of theses
ggplot(emu_theses_words_count |> 
         filter(title == unique(title)[1:6]),
       aes(n/total, fill = title)) +
  geom_histogram(show.legend = FALSE, binwidth = 0.0001) +
  xlim(NA, 0.0009) +
  facet_wrap(~ title, ncol = 3, scales = "free_y")
```

```{r}
# Examine Zipf's law for the EMU theses
## Calculate rank and term frequency
freq_by_rank <- emu_theses_words_count |> 
  group_by(title) |> 
  mutate(rank = row_number(),
         `term frequency` = n/total)

@@

@@

### Word embeddings

[Describe word embeddings ...]

```{r word-embeddings-pmi}
#| include: false

# To what extent do words co-occur with other words?
tidy_pmi <- get_pmi(emu_theses, title_col = "title", text_col = "text_clean")

The high-dimensional sparse matrix of word features is projected into reduced, 100-dimensional set of features.

```{r word-embeddings-svd}

# Determine word vectors using SVD, a method for dimensionality reduction
tidy_word_vectors <- tidy_pmi |> 
  widely_svd(
    item1, item2, pmi,
    nv = 100, maxit = 1000
  )

[What terms should be checked with the function below?]

```{r word-embeddings-nn}

# Calculate nearest neighbours of a given word
tidy_word_vectors |> 
  nearest_neighbors("resilience")

The first [?] components... [what do they show?]

```{r word-embeddings-vis}

# The first n components with top words found in the corpus of theses
visualise_first_n_components(tidy_word_vectors, 24, 12)

```{r doc-embeddings}

# Summarise word embeddings into document embeddings for modeling purposes
word_matrix <- tidy_emu_theses |> 
  count(title, word) |> 
  cast_sparse(title, word, n)

embedding_matrix <- tidy_word_vectors |> 
  cast_sparse(item1, dimension, value)

doc_matrix <- word_matrix %*% embedding_matrix

dim(doc_matrix)

doc_matrix

>>>>

# Load packages ----
## List packages to be loaded
packages <- c(
  "here",                                        # Managing paths
  "tidyverse", "reshape2",                       # Data manipulation and visualisation
  "tidytext", "SnowballC", "tm", "topicmodels",  # Text processing and analysis
  "tidygeocoder",                                # Working with spatial data
  "igraph", "ggraph"                             # Network analysis and visualization
)

@@

Description of text extraction; raw data available on 4TU.ResearchData at...

```{r extract-pdf-text}
#| echo: false
#| cache: true

data_path <- here("analysis", "data", "derived_data", "emu_theses_with_text.csv")

if (!file.exists(data_path)) {
  # On the first run, this function might take extra time as a Python environment needs to be set up
  emu_theses$text <- convert_pdf_text(pdf_paths)
  write_csv(emu_theses, data_path)
} else {
  emu_theses <- read_csv(data_path)
}

@@

@@

### Tokenisation

@@

### Stop words

For analyses in which word-level tokenisation was chosen, stop words, i.e., common words or words that do not add much meaning, were removed in the data preparation stage. Global, subject-specific and document stop words [@hvitfeldt2021a] were differentiated. As all theses were written in English, the pre-made "snowball" lexicon was used to remove global stop words. <!-- We tried to create a custom global stop word list on the entire corpus, by removing high-frequency words, starting with 20 words and increasing by 10 until reaching words that are not appropriate as stop words, but no significant imoprvement was observed in the result. --> Subject-specific stop words, we manually constructed a list based on our domain knowledge. Words that are commonly used to structure a theses, such as "preface", "contents", and "introduction", were added as subject-specific stop words as well. Finally, titles and author names were removed as document stop words on document-level.

@@

emu_theses$text_clean <- emu_theses$text_clean |>
  str_remove_all(find_meta_stopwords(emu_theses)) |> 
  str_remove_all(urbanism_stopwords(
    add_stopwords = c("emu", "european postgraduate master in urbanism"))) |>
  str_remove_all(thesis_stopwords(add_stopwords = c("advisor", "prof")))

emu_theses_words <- emu_theses |> 
  select(title, text_clean) |> 
  unnest_tokens(output = word, input = text_clean) |>   # remove punctuation, convert to lowercase, separate all words
  anti_join(tidytext::stop_words, by = "word")

@@

Additional checks were made to the document in search for previously undetected stop words. The checks included identifying words consisting of four or less characters that extensively used in the text corpus.

```{r}
#| label: short-stop-words

#short_words(emu_theses$text_clean, scope = 'corpus')
short_words(emu_theses_words$word)

@@

Stemming is a pre-processing step that needs to be carefully thought through, as it might either degrade topic modeling or produce no meaningful results [@hvitfeldt2021a].

```{r stemming}

emu_theses_words %>% 
  mutate(stem = wordStem(words = word, 
                         language = "porter")) |>  # Implement the Porter stemming algorithm for English provided by the SnowballC package
  count(stem, sort = TRUE)
```

@@

Similar to clustering on numeric data, topic modeling is a method of unsupervised classification of topics found in a group of documents [@silge2017].

```{r}
# The {topicmodels} package requires a document-term matrix as input
emu_theses_dtm <- convert_to_dtm(emu_theses_words, "title", "word")
emu_theses_dtm

# We will use the Latent Dirichlet Allocation (LDA) algorithm for topic modeling
# In LDA, each document is a mixture of topics and each topic is a mixture of words
emu_theses_lda <- LDA(emu_theses_dtm, k = 2, control = list(seed = 2023))
emu_theses_lda

# Tidy model with beta statistic
emu_theses_topics <- tidy(emu_theses_lda, matrix = "beta")
emu_theses_topics

# The top 10 words that are the most common in each topic
emu_theses_top_terms <- emu_theses_topics |> 
  group_by(topic) |> 
  top_n(20, beta) |> 
  ungroup() |> 
  arrange(topic, -beta)

emu_theses_top_terms |> 
  mutate(term = reorder(term, beta)) |> 
  ggplot(aes(term, beta, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ topic, scales = "free") +
  coord_flip()

# 
beta_spread <- emu_theses_topics |> 
  mutate(topic = paste0("topic", topic)) |> 
  spread(topic, beta) |> 
  filter(topic1 > 0.001, topic2 > .001) |> 
  mutate(log_ratio = log2(topic2/topic1))

beta_spread

beta_spread %>%
  group_by(direction = log_ratio > 0) %>%
  slice_max(abs(log_ratio), n = 20) %>% 
  ungroup() %>%
  mutate(term = reorder(term, log_ratio)) %>%
  ggplot(aes(log_ratio, term)) +
  geom_col() +
  labs(x = "Log2 ratio of beta in topic 2 / topic 1", y = NULL)

emu_theses_documents <- tidy(emu_theses_lda, matrix = "gamma")
emu_theses_documents

emu_theses_documents |> 
  ggplot(aes(x = reorder(str_sub(document, 1, 50), gamma), y = gamma, fill = topic)) + 
  geom_bar(stat = "identity") + 
  coord_flip()

emu_theses_documents |> 
  inner_join(emu_theses |> select(title, graduation_year), by = c("document" = "title")) |> 
  ggplot(aes(x = reorder(str_sub(document, 1, 50), gamma), y = gamma, fill = topic)) + 
  geom_bar(stat = "identity") + 
  coord_flip() + 
  facet_wrap(~ graduation_year, ncol = 1, scales = "free", drop = TRUE)

emu_theses_documents |> 
  inner_join(emu_theses |> select(title, graduation_year), by = c("document" = "title")) |> 
  ggplot(aes(x = factor(graduation_year), y = topic, fill = factor(topic))) +
  geom_col()
```

@@

```{r}
# Which are the top n most frequently used words in each thesis?
get_top_words_per_document(data = emu_theses_words, 
                           top_n = 3, 
                           title_col = "title", 
                           word_col = "word")
```

```{r}
# Which are the top n most frequently used words in all theses`?
get_top_words_per_corpus(data = emu_theses_words, 
                         top_n = 5, 
                         word_col = "word")
```

```{r}
# Which are the top n most frequently used words in all theses`?
emu_theses_top_20 <- get_top_words_per_corpus(data = emu_theses_words, 
                         top_n = 20, 
                         word_col = "word")
```

<!-- ```{r} -->

<!-- # Which theses contain "fabric" in the top 20 most frequently used words? -->

<!-- emu_theses_top_20 %>%  -->

<!--   filter(word == "fabric") -->

<!-- emu_theses_top_20 %>% -->

<!--   slice(1:40) %>%  -->

<!--   ggplot(aes(x = reorder(word, n), y = n, fill = title)) + -->

<!--   geom_col() + -->

<!--   xlab(NULL) + -->

<!--   coord_flip() + -->

<!--   labs(y = "Count", -->

<!--        x = "Unique words", -->

<!--        title = "Count of unique words found in thesis", -->

<!--        subtitle = "Stop words were removed from the list") + -->

<!--   facet_wrap( ~ title, scales = "free", ncol = 2) -->

<!-- ``` -->

@@

## TO DO: generalise bigrams and trigrams into ngrams function

## Bigrams
emu_theses_bigrams <- emu_theses |> 
  unnest_tokens(bigram, text, token = "ngrams", n = 2) |> 
  separate(bigram, into = c("first", "second"), sep = " ", remove = FALSE) |> 
  anti_join(stop_words, by = c("first" = "word")) |> 
  anti_join(stop_words, by = c("second" = "word")) |> 
  filter(str_detect(first, "[a-z]") & 
         str_detect(second, "[a-z]"))

bigram_counts <- emu_theses_bigrams |> 
  count(first, second, sort = TRUE)

bigram_counts

## Trigrams
emu_theses_trigrams <- emu_theses |> 
  unnest_tokens(trigram, text, token = "ngrams", n = 3) |> 
  separate(trigram, into = c("first", "second", "third"), sep = " ", remove = FALSE) |> 
  filter(!first %in% stop_words$word,
         !second %in% stop_words$word,
         !third %in% stop_words$word) |>  
  filter(str_detect(first, "[a-z]"), 
         str_detect(second, "[a-z]"),
         str_detect(third, "[a-z]"))

trigram_counts <- emu_theses_trigrams |> 
  count(first, second, sort = TRUE)

@@

emu_theses_bigrams_top <- emu_theses_bigrams |>
  group_by(title) |>
  count(bigram, sort = TRUE) |>
  slice_max(n, n = 20) |>        ### top_n is superseded; better use slice_max instead
  # filter(n > 50)                ### filter words with a minimum count
  ungroup() |>
  mutate(bigram = reorder(bigram, n))

emu_theses_bigrams_top_all <- emu_theses_bigrams |>
  count(bigram, sort = TRUE) |>
  slice_max(n, n = 20) |>        ### top_n is superseded; better use slice_max instead
  # filter(n > 50)                ### filter words with a minimum count
  mutate(bigram = reorder(bigram, n))

emu_theses_bigrams_top |>
  filter(title == unique(title)[1:6]) |> 
  ggplot(aes(x = bigram, y = n, fill = title)) +
  geom_col() +
  xlab(NULL) +
  coord_flip() +
  labs(y = "Count",
       x = "Unique words",
       title = "Count of unique words found in thesis",
       subtitle = "Stop words were removed from the list") +
  facet_wrap( ~ title, scales = "free", ncol = 2) +
  theme(legend.position = "none") 
```

```{r}
#| label: tf-idf

emu_theses_words

emu_theses_words_count <- emu_theses_words |> 
  count(title, word, sort = TRUE) |> 
  ungroup()

emu_theses_words_total <- emu_theses_words_count |> 
  group_by(title) |> 
  summarise(total = sum(n))

emu_theses_words_count <- left_join(emu_theses_words_count, emu_theses_words_total)

# emu_theses_words_count |> 
#   group_by(title) |> 
#   summarise(total = unique(tot

# Preview term frequency distributions in a subset of theses
ggplot(emu_theses_words_count |> 
         filter(title == unique(title)[1:6]),
       aes(n/total, fill = title)) +
  geom_histogram(show.legend = FALSE, binwidth = 0.0001) +
  xlim(NA, 0.0009) +
  facet_wrap(~ title, ncol = 3, scales = "free_y")

# Examine Zipf's law for the EMU theses
## Calculate rank and term frequency
freq_by_rank <- emu_theses_words_count |> 
  group_by(title) |> 
  mutate(rank = row_number(),
         `term frequency` = n/total)

@@

## Calculate tf-idf
emu_theses_words_count <- emu_theses_words_count |> 
  bind_tf_idf(word, title, n)
emu_theses_words_count

emu_theses_words_count |> 
  select(-total) |> 
  arrange(desc(tf_idf))

@@

The `tf-idf` statistic shows high values for the names of thesis locations that tend to be high-frequency thesis-specific words. As such, it shows words that are the most important to one document in a collection of documents [@silge2017].
